Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-12-12 analysis/paper_draft3.Rmd
edit introduction to add power prior justification

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, establishing a prima facie case, and proving that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)).

@@

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

Here, we demonstrate this approach with simulation studies. Then, we present a software prototype that encodes the same approach with actual historical strike data. With that data, the prototype shows how attorneys and other can this approach to help them decide to argue, and ultimately prove, illegal bias in peremptory-challenge use in real time during jury selection.

>>>>

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, establishing a prima facie case, and proving that bias in the particu-lar case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is the possible incompatibility between current and historical data. This incompatibility could arise due to missing data, or if attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity).[may need fleshing out]

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly incorporate the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posteri-or distribution developed with the historical information and the power prior.

The use of the power prior in this context allows the explicit control of the degree of incompatibility between the current and historical trial data. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected when incompatibility is expected to be high, and higher values (closer to 1) of the power prior parameter selected when incompatibility is low. For now, the choice of alpha is placed in the hands of the user, who will need to be able to defend their choice when presenting a _Batson_ challenge before the court.

Here, we demonstrate this approach with simulation studies where we vary the degree of incompatibility between current and historical trial data, and test our ability to detect bias using different values of alpha. Then, we present a software prototype that encodes the same approach with actual historical strike data. With that data, the prototype shows how attorneys and other can this approach to help them decide to argue, and ultimately prove, illegal bias in peremptory-challenge use in real time during jury selection.
